Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-19 paper.tex
fix errors
Move example figures around

\documentclass[runningheads]{llncs}

@@

For slot~2, unification yields the substitution $σ ≔ \{y ↦ b, z ↦ c\}$, so we insert $h₁$ into $H_{y}(2, b)$.
No matches at level~1 can be extended by $h₁$ since
\[
  \bigcap_{x \in \vars(2)} M_{x}(1, σ(x)) = M_{y}(1, b) = \emptyset
\]
At this point, $H$ and $M$ are in the state shown in Diagram~\eqref{eq:example1:state1}.

Now we add $h₂ : a ≤ b$, which again unifies with both slots.
For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $M_{y}(2, b) = \{h₁\}$.
Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
For slot~2, we add $h₂$ to $H_{y}(2, a)$.
There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state in Diagram~\eqref{eq:example1:state2}.

    \begin{minipage}{0.40\textwidth}
      \begin{center}
      \begin{equation}\label{eq:example1:state1}
        H_y : 
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
              \node at (0,0) {2}
                child {node {$b$}
                  child {node {$\{h₁\}$}}};
        \end{tikzpicture}
        && \ \
        M_y : 
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
              \node at (0, 0) {1}
                child {node {$c$}
                  child {node {$\{[h₁]\}$}}};
        \end{tikzpicture}
      \end{equation}
    \end{center}
    \end{minipage}
    \begin{minipage}{0.55\textwidth}
      \begin{center}
      \begin{equation}\label{eq:example1:state2}
        H_y : 
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, level distance=4em, scale=0.5]
              \node at (0,0) {2}
              child {node[xshift=-2] {$b$}
                child {node {$\{h₁\}$}}}
              child {node[xshift=2] {$a$}
                child {node {$\{h₂\}$}}};
        \end{tikzpicture}
        &&
        M_y : 
        \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
              \node at (-1, 0) {1}
              child {node[xshift=-3] {$c$}
                child {node {$\{[h₁]\}$}}}
              child {node[xshift=1] {$b$}
                child {node {$\{[h₂]\}$}}}
        \end{tikzpicture}
      \end{equation}
      \end{center}
    \end{minipage}

\paragraph{Example 2.}
We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$, which is added to $H$ with slot~2.
There are no matches yet, so there is nothing to extend.
We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the rule state in Diagram~\eqref{eq:example2:state1}:
  \begin{align}\label{eq:example2:state1}
    H_x : 
    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
      \node at (0, 0) {2}
        child {node {$a$}
          child {node {$\{h_1\}$}}};
    \end{tikzpicture}
    &&
    H_y : 
    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
      \node at (0, 0) {2}
        child {node {$a$}
          child {node {$\{h_1\}$}}};
    \end{tikzpicture}
    &&
    M_x : 
    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
      \node at (0, 0) {1}
      child {node {$a$}
        child {node {$\{[h_2]\}$}}};
    \end{tikzpicture}
    &&
    M_y : 
    \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
      \node at (0, 0) {1}
      child {node {$b$}
        child {node {$\{[h_2]\}$}}};
    \end{tikzpicture}
  \end{align}

To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
These are the variables shared between slot~$\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
Hence the compatible hypotheses are
\[
  \bigcap_{x \in \vars(\lvl(m) + 1)} H_{x}(i + 1, \sub(m)(x)) = H_{x}(2, a) \cap H_{y}(2, b) = \emptyset
\]
This means the match cannot currently be extended.
We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$.
This yields the rule state in Diagram~\eqref{eq:example2:state2}.
Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
\[
  \bigcap_{x \in \vars(i)} M_{x}(i - 1, \sub_{i}(x)) = M_{x}(1, a) \cap M_{y}(1, b) = \{[h₂]\}
\]
This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.

\begin{align}\label{eq:example2:state2}
  H_x : 
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
    child {node {$a$}
      child {node {$\{h_1, h_3\}$}}};
  \end{tikzpicture}
  &&
  H_y : 
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
    child {node {$a$}
      child {node {$\{h_1\}$}}}
    child {node {$b$}
      child {node {$\{h_3\}$}}};
  \end{tikzpicture}
  &&
  M_x : 
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
    child {node {$a$}
      child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
  &&
  M_y : 
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
    child {node {$b$}
      child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
\end{align}

>>>>

\documentclass[runningheads,leqno]{llncs}

@@

For slot~2, unification yields the substitution $σ ≔ \{y ↦ b, z ↦ c\}$, so we insert $h₁$ into $H_{y}(2, b)$.
No matches at level~1 can be extended by $h₁$ since
\[
  \bigcap_{x \in \vars(2)} M_{x}(1, σ(x)) = M_{y}(1, b) = \emptyset
\]
At this point, $H$ and $M$ are in state~\eqref{eq:example1:state1}, shown below.

\begin{center}
  \begin{minipage}{0.44\textwidth}
    \begin{center}
    \begin{equation}\label{eq:example1:state1}
      H_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
            \node at (0,0) {2}
              child {node {$b$}
                child {node {$\{h₁\}$}}};
      \end{tikzpicture}
      M_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
            \node at (0, 0) {1}
              child {node {$c$}
                child {node {$\{[h₁]\}$}}};
      \end{tikzpicture}
    \end{equation}
  \end{center}
  \end{minipage}
  \begin{minipage}{0.55\textwidth}
    \begin{center}
    \begin{equation}\label{eq:example1:state2}
      H_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, level distance=4em, scale=0.5]
            \node at (0,0) {2}
            child {node[xshift=-2] {$b$}
              child {node {$\{h₁\}$}}}
            child {node[xshift=2] {$a$}
              child {node {$\{h₂\}$}}};
      \end{tikzpicture}
      M_y\colon
      \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)}, outer sep=auto, level distance=4em, scale=0.5]
            \node at (-1, 0) {1}
            child {node[xshift=-3] {$c$}
              child {node {$\{[h₁]\}$}}}
            child {node[xshift=1] {$b$}
              child {node {$\{[h₂]\}$}}};
      \end{tikzpicture}
    \end{equation}
    \end{center}
  \end{minipage}
\end{center}

Now we add $h₂ : a ≤ b$, which again unifies with both slots.
For slot~1, $\sub₁(h₂)(y) = b$, so we add the match $[h₂]$ to $M_{y}(1, b)$.
We then check for hypotheses in slot~2 that also instantiate $y$ with $b$, and indeed we find $M_{y}(2, b) = \{h₁\}$.
Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : a ≤ c$ to be added later.
For slot~2, we add $h₂$ to $H_{y}(2, a)$.
There are no compatible matches at level~1 so we are done with $h₂$, yielding the final rule state~\eqref{eq:example1:state2}.

\paragraph{Example 2.}
We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$, which is added to $H$ with slot~2.
There are no matches yet, so there is nothing to extend.
We then add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$, yielding the following rule state:
\begin{align*}
  H_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}};
  \end{tikzpicture}
  &&
  H_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
      child {node {$a$}
        child {node {$\{h_1\}$}}};
  \end{tikzpicture}
  &&
  M_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
    child {node {$a$}
      child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
  &&
  M_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
    child {node {$b$}
      child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
\end{align*}

To find hypotheses compatible with $m$, we consider the set of variables $\vars(\lvl(m) + 1)$.
These are the variables shared between slot~$\lvl(m) + 1 = 2$ and all previous slots, i.e.\ $x$ and $y$.
Hence the compatible hypotheses are
\[
  \bigcap_{x \in \vars(\lvl(m) + 1)} H_{x}(i + 1, \sub(m)(x)) = H_{x}(2, a) \cap H_{y}(2, b) = \emptyset
\]
This means the match cannot currently be extended.
We now turn to the remaining hypothesis $h_3 : B~a~b$, adding it first to $H$.
This yields the following rule state:
\begin{align*}
  H_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
    child {node {$a$}
      child {node {$\{h_1, h_3\}$}}};
  \end{tikzpicture}
  &&
  H_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {2}
    child {node {$a$}
      child {node {$\{h_1\}$}}}
    child {node {$b$}
      child {node {$\{h_3\}$}}};
  \end{tikzpicture}
  &&
  M_x\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
    child {node {$a$}
      child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
  &&
  M_y\colon
  \begin{tikzpicture}[baseline={([yshift=-.5ex]current bounding box.center)},outer sep=auto, level distance=4em, scale=0.50]
    \node at (0, 0) {1}
    child {node {$b$}
      child {node {$\{[h_2]\}$}}};
  \end{tikzpicture}
\end{align*}
Since $h₃$ matches slot $i = 2$, the matches that can be extended by $h₃$ are
\[
  \bigcap_{x \in \vars(i)} M_{x}(i - 1, \sub_{i}(x)) = M_{x}(1, a) \cap M_{y}(1, b) = \{[h₂]\}
\]
This produces the complete match $[h₂, h₃]$, so we can add a new hypothesis of type $C~a~b$.
